---
title: "Meta-analysis of synesthetic metaphor tables"
author: "Bodo"
date: "2023-07-10"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script presents the results of an ongoing meta-analysis of synesthetic metaphors, expressions such as the adjective-noun pair *sweet melody* which combines a taste word (*sweet*) with a sound-related word (*melody*).

A number of people have performed corpus analysis counting how many crossmodal expressions fit a proposed "hierarchy of the senses": touch < taste < smell < sound < sight

In this meta-analysis, we compiled all sorts of tables listing token counts and then counted the proportion of "hierarchy-consistent" cases, i.e., how many of the observed tokens of synesthetic metaphors are consistent with the hierarchy? We will look at this, and then see how much particular mappings, especially touch->sound and sight->sound may matter.

What we want to show is the distribution of the proportion of hierarchy-consistent cases across studies, and then again the same without the cells in question. We will also want to plot the results from a logistic regression analysis on top of these distributions.

# Setup

```{r warning = FALSE, message = FALSE}
library(tidyverse)     # for data processing and visualization
library(brms)          # for fitting bayesian models
library(patchwork)     # for multi-plot arrays
```

Load the data:

```{r warning = FALSE, message = FALSE}
syn <- read_csv('../data/winter_strik_lievers_synesthesia_meta_analysis.csv')
```

Load the pre-compiled models. The model `proportion_model` is a logistic regression fitted on the proportions including touch->sound and sight->sound, the model `prop_no_TS_no_VS_mdl` is the proportion without these cells.

```{r}
load('../models/proportion_model.RData')
load('../models/prop_no_TS_no_VS_mdl.RData')
```

# Calculate missing statistics

Let's look at the data together:

```{r}
syn
```

The column `proportion` is simply the `hierarchy_tokens` column divided by the `total` column. We need to compute the missing proportion of hierarchy-consistent tokens that exclude touch->sound and sight->sound, for which we use the column called `hierarchy_tokens_noTS_noVS`. Let's call the resultant proportions `prop_red` for "proportion reduced".

```{r}
syn <- mutate(syn,
              prop_red = hierarchy_tokens_noTS_noVS / total)
```

# Density plot of proportions

Make a plot of this, which we'll save in an object called `prop_p`. We'll add the 95% credible interval of the model's estimate for the proportion of hierarchy-consistent cases into the plot as well. This way people can see the raw data and the model.

The informationa bout the model estimates comes from applying the `fixef()` function (for extracting fixed effects) to the logistic regression model:

```{r}
fixefs <- fixef(prop_mdl)
fixefs_red <- fixef(prop_no_TS_no_VS_mdl)

# Show:

fixefs
```

Then we build the plot. We'll plot two separate geoms with the density curve, each time drawing from a different column. Then we use `geom_vline()` to add a vertical line for chance, which constitutes 11/20 in this case (for reasons that we don't need to get into here). We use the `annotate()` function profusely. This function can plot all sorts of extra elements for geoms specified in the first argument (below: `segment`, `point`, `text`, `label`). Notice the difference between `text` and `label` in the plot below and compare it to the outcome of the plot. To plot the coordinates for the 95% credible intervals, we use the content of the `fixefs` object (from model including all cells) and the `fixefs_red` object (from model excluding touch->sound and sight->sound).

Important technical detail that goes beyond the content of this workshop: we need to wrap `plogis()` around the estimates from the model because the model reports all values in log odds, otherwise known as "logits". The `plogis()` function applies the logistic regression function to these logits so that they can be plotted on the more interpretable proportion scale.

```{r warning = FALSE, message = FALSE, fig.width = 6.5, fig.height = 4}
# Core of plot:

prop_p <- syn %>% 
  ggplot() +
  geom_density(mapping = aes(x = proportion),
               fill = 'purple', alpha = 0.8) +
  geom_density(mapping = aes(x = prop_red),
               fill = 'pink', alpha = 0.8) +
  geom_vline(xintercept = 0.55, linetype = 'dashed')

# Annotations:

prop_p <- prop_p +
  annotate('segment',
           x = plogis(fixefs[, 'Q2.5']),
           xend = plogis(fixefs[, 'Q97.5']),
           y = 0.4, yend = 0.4,
           arrow = arrow(ends = 'both', angle = 90, length = unit(.2, 'cm'))) +
  annotate('point',
           x = plogis(fixefs[, 'Estimate']),
           y = 0.4, pch = 15, size = 3) +
  annotate('segment',
           x = plogis(fixefs_red[, 'Q2.5']),
           xend = plogis(fixefs_red[, 'Q97.5']),
           y = 0.4, yend = 0.4,
           arrow = arrow(ends = 'both', angle = 90, length = unit(.2, 'cm'))) +
  annotate('point',
           x = plogis(fixefs_red[, 'Estimate']),
           y = 0.4, pch = 15, size = 3)
  

# Add annotations, axes, labels to plot:

prop_p <- prop_p +
  annotate('text',
           x = 0.555,
           y = 9,
           label = 'Chance:\n55% = 11/20 cells',
           hjust = 0) +
  annotate('label',
           x = 0.45,
           y = 2.1,
           fill = 'pink',
           label = 'Without:\ntouch->sound\nsight->sound',
           hjust = 1) +
  xlab('Proportion of hierarchy consistent cases') +
  ylab('Density') +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 10)) +
  scale_x_continuous(limits = c(0.2, 1),
                     breaks = seq(0.2, 1, 0.1))

# Cosmetic tweaking:
  
prop_p <- prop_p +
  theme_classic() +
  theme(axis.title = element_text(face = 'bold',
                                  size = 12),
        axis.title.x = element_text(margin = margin(t = 12)),
        axis.title.y = element_text(margin = margin(r = 8)))

# Show plot in markdown and also save externally outside of R:

prop_p
ggsave(plot = prop_p,
       filename = '../figures/png/synesthesia_meta_analysis_average_proportion.png',
       width = 6.5, height = 4)
ggsave(plot = prop_p,
       filename = '../figures/pdf/synesthesia_meta_analysis_average_proportion.pdf',
       width = 6.5, height = 4)
```

This completes this analysis.
